Appendix 3: Estimating species interactions

### 2015-11-02 Harris_2015/manuscript-materials/draft.tex
etc

\textbf{Title:} Estimating species interactions from co-occurrence data
with Markov networks

@@

\textbf{Abstract:} Estimating species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can also be estimated
from observed co-occurrence rates via maximum likelihood, controlling
for indirect effects. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods,
correctly isolating direct interactions between species pairs even when
indirect interactions or abiotic factors largely overpowered them. Two
computationally efficient approximations, based on controlling for
indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = -2\), then each species' odds of
occurrence are \(exp{2}\) times lower when the other one is present (as
compared with otherwise comparable sites). Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect). Of
course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' overall co-occurrence patterns can be more
complicated, and may require summing over the different possible
assemblages (Figure 2B) to determine the overall outcomes.

@@

\noindent
\textbf{Twenty species.} Despite some variability across contexts
(Figure 3), the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) outperformed those that did not. Of these,
the Markov network consistently performed the best. As anticipated by
Lee and Hastie (2012), the Markov network produced very similar
estimates to the generalized linear models (Figure 4A), but with higher
accuracy---especially with smaller data sets. As reviewed in Gotelli and
Ulrich (2009), most analyses in this field in ecology involve fewer than
50 sites, so this is the most important regime for model performance. As
shown in Appendix {[}{[}X{]}{]}, the standard errors associated with the
estimates in Figure 3 are small (less than 0.01), so these differences
should not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). The estimates
from these three approaches were tightly correlated, after controlling
for the size of the landscape, suggesting that the null model only
contains a noisy version of the same information that could be obtained
more easily and interpretably with simple correlation coefficients
(Figure 4B).

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's Type I error rate (defined
here as its probability of rejecting the null hypothesis for a pair of
species where \(|\beta_{ij}|<0.1\) depended on the simulation type: 0.03
for simulations that matched the model's assumptions, versus 0.15 for
simulations that included environmental heterogeneity
{[}{[}Appendix{]}{]}. For the null model, these rates were 0.31 and
0.50, respectively. Figure 4C shows, across a range of true interaction
strengths, the probability that the null model or the Markov network
will predict the wrong sign of the interaction with 95\% confidence. The
null model makes such errors nearly 8 times as often as the Markov
network, even though it only reject the null hypothesis twice as often
overall {[}{[}Appendix{]}{]}. The Markov network's errors were also more
concentrated around 0, so it never misclassified strong interactions
like the null model did.

@@

\noindent
The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where environmental heterogeneity and indirect interactions cause
ecologists' typical null modeling approaches to reliably fail. Partial
covariances and generalized linear models can both provide
computationally efficient approximations, but with somewhat lower
accuracy, especially for typically-sized data sets with small numbers of
sites (Gotelli and Ulrich 2009). The difference in accuracy may be
larger for real data sets than for the simulated landscapes in Figure 3,
however; linear approximations to the Markov network make larger errors
when the interaction matrix is structured (e.g.~due to guilds or trophic
levels; Loh and Wainwright 2013). Similarly, the separate generalized
linear models for each species can severely overfit in some cases (Lee
and Hastie 2012). The full Markov network should thus be preferred to
the approximations when it is computationally feasible.

Compositional data only contains enough degrees of freedom to estimate
one interaction per species pair, so none of these methods can identify
the exact nature of the pairwise interactions (e.g.~which species in a
positively-associated pair is facilitating the other; cf Schmidt and
Murphy 2012). To estimate asymmetric interactions, such as commensalism
or predation, ecologists could use time series, behavioral observations,
manipulative experiments, or natural history. These other sources of
information could also be used to augment the likelihood function with
an informative prior distribution, reducing ecologists' error and
uncertainty.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that predicts the
consequences of species interactions (as opposed to null simulations
that omit interactions entirely).

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models are well-understood. Using the same framework for species
interactions would thus allow ecologists to tap into a vast set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would help incorporate networks of biotic interactions into
species distribution models (Pollock et al. 2014) and lead to a better
understanding of the interplay between biotic and abiotic effects on
community structure. Alternatively, models could allow one species to
alter the relationship between two other species (Tjelmeland and Besag
1998, cf Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking (Harris et al. 2011, Xiao et al. 2015), but deviations from a
particular null model must be interpreted with care (Roughgarden 1983).
Even in small networks with three species, it may simply not be possible
to implicate specific ecological processes like competition by rejecting
a general-purpose null (Gotelli and Ulrich 2009), especially when the
test statistic is effectively just a correlation coefficient Figure 4A.
When a non-null backdrop is not controlled for, Type I error rates can
skyrocket, the apparent sign of the interaction can change, and null
models can lead to reliably incorrect inferences (Figure 1, Figure 4).

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the
extent that the models' relative performance on real data sets is
similar to the range of results shown in Figure 3A, scientists in this
field could more than double their explanatory power by switching from
null models to Markov networks (or increase it nearly as much with
linear or generalized linear approximations). Regardless of the methods
ecologists ultimately choose, controlling for indirect effects could
clearly improve our understanding of species' direct effects on one
another and on community structure.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing
species. The tree (top) tends not to co-occur with either of the two
shrub species, as indicated by the strongly negative coefficient linking
them. The two shrub species also compete with one another, but more
weakly (circled coefficient). \textbf{B.} In spite of the competitive
interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors
positively correlated (circled). \textbf{C.} Controlling for the tree
species' presence with a conditional (all-else-equal) approach such as a
partial covariance or a Markov network leads to correct identification
of the negative shrub-shrub interaction (circled).

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its
\(\alpha\) and \(\beta\) values. The abiotic environment favors the
occurrence of each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta_12\) coefficient is
consistent with competition between the two species. \textbf{B.} The
coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
and includes \(\beta_12\) only when both are present (\(y_1y_2 = 1\)).
The normalizing constant \(Z\), ensures that the four probabilities sum
to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
frequencies of all possible co-occurrence patterns between the two
species of interest, as calculated in the previous panel. \textbf{D.} If
\(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
the same resources), then the each species' overall occurrence rates
would increase.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled
locations.

\noindent \textbf{Figure 4: A.} The Markov network's estimated
interaction coefficients were generally very similar to the GLM
estimates. \textbf{B.} The null model's estimates typically matched the
correlation coefficient, after controlling for landscape size.
\textbf{C.} For any given interaction strength, the null model was much
more likely to misclassify its sign with 95\% confidence than the Markov
network was.

>>>>

\textbf{Title:} Inferring species interactions from co-occurrence data
with Markov networks

@@

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can also be estimated
from observed co-occurrence rates via maximum likelihood, controlling
for indirect effects. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods,
correctly isolating direct interactions between species pairs even when
indirect interactions or abiotic factors largely overpowered them. Two
computationally efficient approximations, based on controlling for
indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = +2\), then each species' odds of
occurrence would be \(e^2\) times higher when the other one is present
(as compared with otherwise equivalent sites). Because the relative
probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else
equal---to produce assemblages that have many pairs of
positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect). Of
course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then predicting species' overall co-occurrence rates can be
more complicated, and may require summing over the different possible
assemblages (Figure 2B).

@@

\noindent
\textbf{Twenty species.} Despite some variability across contexts
(Figure 3), the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) outperformed those that did not. Of these,
the Markov network consistently performed the best. As anticipated by
Lee and Hastie (2012), generalized linear models closely approximated
the Markov network estimates, especially when the data sets were very
large. As reviewed in Gotelli and Ulrich (2009), however, most analyses
in this field involve fewer than 50 sites, so model performance on the
smaller landscapes where the gap was larger is most important. As shown
in Appendix {[}{[}X{]}{]}, the standard errors associated with the
estimates in Figure 3 are small (less than 0.01), so the differences
among methods should not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). The estimates
from these approaches were tightly correlated (after controlling for the
size of the landscape) suggesting that the null model only contains a
noisy version of the same information that could be obtained more easily
and interpretably with simple correlation coefficients (Figure 4B).

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as its probability of rejecting the null hypothesis
for a pair of species where \(|\beta_{ij}|<0.1\)) depended on the
simulation type: 0.03 for simulations that matched the model's
assumptions, versus 0.15 for simulations that included environmental
heterogeneity {[}{[}Appendix{]}{]}. For the null model, these rates were
0.31 and 0.50, respectively---far higher than the nominal 0.05 rate.
Figure 4C shows, across a range of true interaction strengths, the
probability that the null model or the Markov network will predict the
wrong sign of the interaction with 95\% confidence. The null model makes
such errors nearly 8 times as often as the Markov network, even though
it only reject the null hypothesis twice as often overall
{[}{[}Appendix{]}{]}. The Markov network's errors were also more
concentrated around 0, so it never misclassified strong interactions
like the null model did.

@@

\noindent
The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where environmental heterogeneity and indirect interactions cause
ecologists' typical null modeling approaches to reliably fail. Partial
covariances and generalized linear models can both provide
computationally efficient approximations, but with somewhat lower
accuracy, especially for typically-sized data sets with small numbers of
sites (Gotelli and Ulrich 2009). The difference in accuracy may be
larger for real data sets than for the simulated landscapes in Figure 3,
however; linear approximations to the Markov network make larger errors
when the interaction matrix is structured (e.g.~due to guilds or trophic
levels; Loh and Wainwright 2013). Similarly, the separate generalized
linear models for each species can severely overfit in some cases (Lee
and Hastie 2012). The full Markov network should thus be preferred to
the approximations when it is computationally tractable.

Compositional data only contains enough degrees of freedom to estimate
one interaction per species pair (Schmidt and Murphy 2012), so none of
these methods can identify the exact nature of the pairwise interactions
(e.g.~which species in a positively-associated pair is facilitating the
other). To estimate asymmetric interactions, such as commensalism or
predation, ecologists could use time series, behavioral observations,
manipulative experiments, or natural history. These other sources of
information could also be used to augment the likelihood function with
an informative prior distribution, reducing ecologists' error and
uncertainty relative to Figure 3's results.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they make many fewer
errors than existing approaches, and can make precise statements about
the conditions where indirect interactions will overwhelm direct ones.
They also provide a simple answer to the question of how competition
should affect a species' overall prevalence, which has important
implications for community-level modeling (Strong et al. 1984).
Specifically, Equation 1 can be used to calculate the expected
prevalence of a species in the absence of biotic influences as
\(e^\alpha/(e^{0} + e^\alpha)\). Competition's effect on prevalence can
then be estimated by comparing this value with the observed prevalence
(e.g.~comparing Figure 2D with Figure 2C). This novel quantitative
result conflicts with most of our null models, which unreasonably assume
that prevalence would be the exactly same in the absence of competition
as it is in the observed data (Roughgarden 1983).

Markov networks---particularly the Ising model for binary networks---are
very well understood, having been studied for nearly a century (Cipra
1987). Tapping into this framework would thus allow ecologists to take
advantage of into a vast set of existing discoveries and techniques for
dealing with indirect effects, stability, and alternative stable states.
Numerous extensions to the basic network are possible as well. For
example, the states of the interaction network can be modeled as a
function of the local abiotic environment (Lee and Hastie 2012), which
would help incorporate networks of biotic interactions into species
distribution models (Pollock et al. 2014) and lead to a better
understanding of the interplay between biotic and abiotic effects on
community structure. Alternatively, models could allow one species to
alter the relationship between two other species (Tjelmeland and Besag
1998, cf Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking (Harris et al. 2011, Xiao et al. 2015), but deviations from a
particular null model must be interpreted with care (Roughgarden 1983).
Even in small networks with three species, it may simply not be possible
to implicate specific ecological processes like competition by rejecting
a general-purpose null (Gotelli and Ulrich 2009), especially when the
test statistic is effectively just a correlation coefficient (Figure
4B). When the non-null backdrop is not controlled for, Type I error
rates can skyrocket, the apparent sign of the interaction can change,
and null models can routinely produce misleading inferences (Figure 1,
Figure 4C, Gotelli and Ulrich (2009)).

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the
extent that the models' relative performance on real data sets is
similar to the range of results shown in Figure 3, scientists in this
field could more than double their explanatory power by switching from
null models to Markov networks (or increase it nearly as much with
linear or generalized linear approximations). Regardless of the methods
ecologists ultimately choose, controlling for indirect effects could
clearly improve our understanding of species' direct effects on one
another and on community structure.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing
species. The tree (top) tends not to co-occur with either of the two
shrub species, as indicated by the strongly negative coefficient linking
them. The two shrub species also compete with one another, but more
weakly (circled coefficient). \textbf{B.} In spite of the competitive
interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors
positively correlated (circled). \textbf{C.} Controlling for trees with
a conditional (all-else-equal) approach such as a partial covariance or
a Markov network leads to correct identification of the negative
shrub-shrub interaction (circled).

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its
\(\alpha\) and \(\beta\) values. The abiotic environment favors the
occurrence of each species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
consistent with competition between the two species. \textbf{B.} The
coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
The normalizing constant \(Z\), ensures that the four probabilities sum
to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
frequencies of all possible co-occurrence patterns between the two
species of interest, as calculated in the previous panel. \textbf{D.}
Without competition (i.e.~with \(\beta_{12}=0\), each species would
occur more often.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations
across the three simulation types.

\noindent \textbf{Figure 4: A.} The Markov network's estimated
interaction coefficients were generally very similar to the GLM
estimates. \textbf{B.} The null model's estimates typically matched the
(negative) correlation coefficient, after controlling for landscape
size. \textbf{C.} For any given interaction strength, the null model was
much more likely to misclassify its sign with 95\% confidence than the
Markov network was.
